Exploratory Data Analysis of Red Wine Quality by John Ingles
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
rw_data<-read.csv("wineQualityReds.csv")

```

# Introduction

In this report, I'll be performing an exploratory data analysis on a data set of
red wine samples collected by Cortez et al.[1].  The data set contains 1,559 red
wine samples with 11 variables on the chemical properties of each.  The quality
of each sample was also recorded by averaging the scores of at least 3 wine
experts.  They provided a rating between 0 (very bad) and 10 (very excellent).
The goal of this analysis is to determine if any of the factors measured
influence the quality of the wines as measured by the experts.

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, str}
str(rw_data)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, summary}
summary(rw_data)
```

A summary of the data set is shown above.  It shows the name, data
type, and quantile distribution of each variable.

I'll look at the results first.

```{r echo=FALSE, message=FALSE, warning=FALSE, quality}
ggplot(data=rw_data, aes(x=quality))+
  geom_bar()
```

The quality distribution is uni-modal around 5 and 6 with a skew towards 6.
Quality values of 0, 1, 2, 9, and 10 are absent.

Next, I'll look at distributions of the independent variables.  I'll start with
the acidity values.

```{r echo=FALSE, message=FALSE, warning=FALSE, fixed.acidity_volatile.acidity}
ggplot(data=rw_data, aes(x=fixed.acidity))+
  geom_histogram(binwidth = 0.1)+
  scale_x_continuous(breaks=seq(4,12,0.5), limits=c(4,12))

ggplot(data=rw_data, aes(x=volatile.acidity))+
  geom_histogram(binwidth = 0.01)+
  scale_x_continuous(breaks=seq(0.2,0.8,0.05), limits=c(0.2,0.8))
```

Fixed acidity is right skewed while volatile acidity is more centered and
slightly bimodal. Fixed acidity values around 7 g/L are most numerous, while
volatile acidity values are more spread out and centered around 0.5 g/L.  Both
values have a few high valued outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, citric.acid}
ggplot(data=rw_data, aes(x=citric.acid))+
  geom_histogram(binwidth = 0.01)+
  scale_x_continuous(breaks=seq(0,0.75,0.05),limits=c(-0.01,0.75))
```

Over 150 of the wine samples had 0 g/L of citric acid followed by spikes of
50, 51, and 68 for samples with 0.02 g/L, 0.24 g/L, and 0.49 g/L respectively.
The other values have counts under 25 for the most part.

```{r echo=FALSE, message=FALSE, warning=FALSE, residual.sugar}
ggplot(data=rw_data, aes(x=residual.sugar))+
  geom_histogram(binwidth = 0.1)

ggplot(data=rw_data, aes(x=residual.sugar))+
  geom_histogram(binwidth = 0.1)+
  scale_x_continuous(limits=c(1,4), breaks=seq(1,4,0.2))
```

After zooming in past the outliers we see a normal distribution centered around
2 g/L with a width of about 1 g/L for the residual sugar counts.

```{r echo=FALSE, message=FALSE, warning=FALSE, chlorides}
ggplot(data=rw_data, aes(x=chlorides))+
  geom_histogram(binwidth = 0.001)+
  scale_x_continuous(limits=c(0.025,0.15), breaks=seq(0.025,0.15,0.01))
```

Chlorides show a similar distribution around 0.75 g/L but with a more narrow
width.

```{r echo=FALSE, message=FALSE, warning=FALSE, sulfur.dioxide}
ggplot(data=rw_data, aes(x=free.sulfur.dioxide))+
  geom_histogram(binwidth = 1)+
  scale_x_continuous(limits=c(0,40),breaks=seq(0,40,5))

ggplot(data=rw_data, aes(x=total.sulfur.dioxide))+
  geom_histogram(binwidth = 1)+
  scale_x_continuous(limits=c(0,150),breaks=seq(0,150,10))
```

The free and total sulfur dioxides are both right skewed distributions with
peaks around 5 mg/L and 25 mg/L respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE, density}
ggplot(data=rw_data, aes(x=density))+
  geom_histogram(binwidth = 0.001)+
  scale_x_continuous(breaks=seq(0.990,1.005,0.001))

summary(rw_data$density)
```

The density of the wines are normally distributed around a mean of 0.9967 with a
width of about 0.0025 g/mL.

```{r echo=FALSE, message=FALSE, warning=FALSE, pH}
ggplot(data=rw_data, aes(x=pH))+
  geom_histogram(binwidth = 0.01)+
  scale_x_continuous(limits=c(2.9,3.75),breaks=seq(2.9,3.75,0.05))

summary(rw_data$pH)
```

Ph is also normally distributed for the most part with a mean of 3.311 and a
width of about 0.25.

```{r echo=FALSE, message=FALSE, warning=FALSE, sulphates}
ggplot(data=rw_data, aes(x=sulphates))+
  geom_histogram(binwidth = 0.01)+
  scale_x_continuous(limits=c(0.35,1.0),breaks=seq(0.35,1.0,0.05))
```

The sulphates measurements are slightly right skewed with a peak at about
0.6 g/L and numerous outliers to the right.

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol}
ggplot(data=rw_data, aes(x=alcohol))+
  geom_histogram(binwidth = 0.1)+
  scale_x_continuous(breaks=seq(8,15,0.5))
```

Finally, The alcohol percentages of the samples are right skewed with a peak at
about 9.5% by volume.

### What is the structure of your dataset?

There are 1599 wine samples in the data set with 11 variables

1. fixed acidity (tartaric acid - g / dm^3)
2. volatile acidity (acetic acid - g / dm^3)
3. citric acid (g / dm^3)
4. residual sugar (g / dm^3)
5. chlorides (sodium chloride - g / dm^3
6. free sulfur dioxide (mg / dm^3)
7. total sulfur dioxide (mg / dm^3)
8. density (g / cm^3)
9. pH
10. sulphates (potassium sulphate - g / dm3)
11. alcohol (% by volume)

### What is/are the main feature(s) of interest in your dataset?

The main feature in the data set is quality rating.  It's my aim to determine
which factors are best for predicting the quality of the wine.  Based on the
info sheet for the source of the data, I suspect volatile acidity, citric acid, 
sulphates, and alcohol content will have the most effect on taste.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Other features which I suspect could possibly affect the quality of the wine
are free sulfur dioxide, ph, and residual sugar.

### Did you create any new variables from existing variables in the dataset?

No new variables were created.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I rescaled the plot for residual sugar since its distribution was narrow with
many outliers.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, CorMatrix}
cor_data <- subset(rw_data, select=-c(X))
colnames(cor_data) <- c("fa", "va","ca","rs","chl","fsd","tsd","den","ph","sul","alc","qual")
round(cor(cor_data),2)
```

Citric acid and fixed acid have a correlation of 0.67, likely owing to citric
acid itself being an acid.  Fixed acid also appears to correlate
positively with density and negatively with ph.  Free sulfur dioxide and total
sulfur dioxide also have a correlation of 0.67, which makes sense since they are 
recording different measures of the same chemical. As far as features directly
correlating with quality, the highest were volatile acidity with -0.39 and
alcohol content with 0.48.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots1}
ggplot(data=rw_data, aes(y=free.sulfur.dioxide, x=total.sulfur.dioxide))+
  geom_point(alpha=1/10)+
  scale_x_continuous(limits=c(0,150),breaks=seq(0,150,10))
```

The ratio of free to total sulfur dioxide is strictly under 0.80 and there
appears to be a slight linear relationship between them as expected.

```{r echo=FALSE, message=FALSE, warning=FALSE, ca_fa}
ggplot(data=rw_data, aes(x=citric.acid, y=fixed.acidity))+
  geom_point(alpha=1/3)+
  geom_smooth(method = 'lm', color='red')+
  scale_x_continuous(limits=c(0,0.75),breaks=seq(0,0.75,0.05))
```

The relationship between citric acid and fixed acidity is decidedly more linear.

```{r echo=FALSE, message=FALSE, warning=FALSE, den_fa}
ggplot(data=rw_data, aes(x=density, y=fixed.acidity))+
  geom_point(alpha=1/3)+
  geom_smooth()+
  scale_x_continuous(limits=c(0.9925,1.0025),breaks=seq(0.9925,1.0025,0.001))

ggplot(data=rw_data, aes(x=pH, y=fixed.acidity))+
  geom_point(alpha=1/3)+
  geom_smooth()+
  scale_x_continuous(limits=c(3,3.75),breaks=seq(3,3.75,0.05))
```

Density and ph also show a relationship to fixed acidity, with density generally
being higher for higher fixed acidity and ph being lower as expected.

Finally, I'll use box plots to get an idea of the effect on quality for each
of the supporting features.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots2}
ggplot(data=rw_data, aes(x=factor(quality), y=volatile.acidity))+
  geom_boxplot()
```
```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots3}
by(rw_data$volatile.acidity, rw_data$quality, summary)
```
There is a clear trend showing an increase in quality as the level of volatile
acidity (acetic acid) decreases.  The mean starts at 0.8840 g/L for samples with
quality 3 and decreases to 0.3700 g/L for samples with quality of 7 or 8.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots4}
ggplot(data=rw_data, aes(x=factor(quality), y=citric.acid))+
  geom_boxplot()
```
```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots5}
by(rw_data$citric.acid, rw_data$quality, summary)
```

The IQR is pretty wide for several quality values such as 3 and 6, but just
looking at the medians and means we can still see a trend of increasing quality
for increasing levels of citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots6}
ggplot(data=rw_data, aes(x=factor(quality), y=sulphates))+
  geom_boxplot()
```
```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots7}
by(rw_data$sulphates, rw_data$quality, summary)
```

The effect of sulphate additives on the quality of the wine samples seems about
the same as that of citric acid, but the IQR's are much smaller.  This points to
the effect of sulphates on quality as being more certain than that citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots8}
ggplot(data=rw_data, aes(x=factor(quality), y=chlorides))+
  geom_boxplot()+
  coord_cartesian(ylim = c(0,0.2))
```

I had to scale this one down because there were many high chloride outliers.
Even after this, there's only a slight difference in quality as the level of
chlorides decreases.


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots9}
ggplot(data=rw_data, aes(x=factor(quality), y=alcohol))+
  geom_boxplot()
```
```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots10}
by(rw_data$alcohol, rw_data$quality, summary)
```

Low levels of alcohol around 10% by volume seem equally distributed at quality
levels of 3, 4 and 5.  After that, there is an increasing trend in quality for
samples with higher levels of alcohol with levels of around 12% being judged to
have the highest quality.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the data set?

There is a relatively high correlation between citric acid and fixed acidity of
0.67.  While there are no correlations over 0.60 for volatile acidity or alcohol.

A plot of citric acid vs. fixed acidity shows what appears to be a linear
relationship, although it's hard to tell for sure with the relatively low number
of data points.

Box plots of volatile acidity, citric acid, sulphates, and alcohol across
quality show a positive influence on quality for alcohol, citric acid and
sulphates, and a negative influence for volatile acidity.

In addition, the box plot of chlorides across quality shows a slight
relationship.  Along with the correlation between quality and chlorides of -0.13,
I think there is enough evidence to consider the effect of chlorides on the
quality along with the main features.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I observed several correlations between the other features:

| Variables        | Correlation           |
| ------------- |:--------------|
| free.sulfur.dioxide vs. total.sulfur.dioxide | 0.67 |
| density vs. fixed.acidity      |   0.67 |
| pH vs. fixed.acidity |    -0.68 |

Plots of these also showed what appeared to be linear relationships.

### What was the strongest relationship you found?

I judged the strongest relationship to be the one between alcohol and
quality based on the correlation of 0.48, along with the trend shown in the set
of box plots of the two variables.

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots1}
ggplot(data=rw_data, aes(x=volatile.acidity, y=alcohol))+
  geom_point(aes(colour=factor(quality)))+scale_color_brewer(palette='RdYlGn')+
  scale_x_continuous(limits=c(0.2,1.2),breaks=seq(0.2,1.2,0.1))
```

With the trends from the previous section, I decided to create a plot of alcohol
versus volatile acidity using a diverging color palette to color the quality of
each sample.  It is clear that the better judged samples have higher alcohol
content and lower volatile acidity.  The samples in the mid-range are mostly in 
the region of low volatile acidity and low alcohol content.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots2}
ggplot(data=rw_data, aes(x=sulphates, y=volatile.acidity))+
  geom_point(aes(colour=factor(quality)))+scale_color_brewer(palette='RdYlGn')+
  scale_x_continuous(limits=c(0.375,1.0),breaks=seq(0.4,1.0,0.05))
```

I then did the same with sulphates instead of alcohol and found a similar result.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots3}
ggplot(data=rw_data, aes(x=volatile.acidity/sulphates, y=alcohol))+
  geom_point(aes(colour=factor(quality)))+scale_color_brewer(palette='RdYlGn')+
  scale_x_log10(breaks=seq(0.15,4.00,0.5))
```

Combining volatile acidity and sulphates with a ratio, applying a log
function, and plotting it against alcohol results in a graph that looks to be
the beginnings of a predictive model for quality.  The divides between good and
bad samples are still clear which means the factors aren't interfering with each
other.

The quality results are ordinal so I will use an Ordinal Logistic Regression to
build a model based on the features.

```{r echo=FALSE, message=FALSE, warning=FALSE, logit}
library(MASS)
rw_data$quality.f<-factor(rw_data$quality)
m1<-polr(quality.f ~ alcohol+volatile.acidity+sulphates,data=rw_data, Hess=TRUE)
summary(m1)
m1.coef<-data.frame(coef(summary(m1)))
m1.coef$pval = round((pnorm(abs(m1.coef$t.value),lower.tail=FALSE)*2),2)
m1.coef

m1.or=exp(coef(m1))
m1.or
```

After running an Ordinal Logistic Regression with alcohol, volatile acidity, and
sulphates as predictors, I get a model that explains the data with high
significance as shown by the p-values shown at 0.  I tried adding other
variables to the model, but most caused no significant change to the Residual
Deviance or AIC.  Adding citric acid caused a slight improvement, but unexpectedly
it had a fractional odds ratio indicating a negative correlation with quality.
Since this went against the previous box plot of citric acid versus quality and
the p value was around 0.20, I dropped it from the model.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

From the plots, we can see the influence of volatile acidity, alcohol, and
sulphates on the quality of the wine samples.  The features don't appear to
influence each other in any direct way.  There was evidence in the previous
sections that citric acid affected fixed acidity and fixed acidity affected pH, 
but none of these variables appeared to have any significant effect on quality.

### Were there any interesting or surprising interactions between features?

In an unused logit model, citric acid had a negative effect on quality, even
though a box plot of citric acid across quality showed a slight positive effect.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Yes, I created a model using Ordinal Logistic Regression with alcohol percentage,
volatile acidity levels, and sulphate levels as predictors.

This test shows that there is a significant lack of evidence to support the null
hypothesis.  This is shown by low p-values for each predictor and a residual
deviance of 3138.035 for a system with 1599 degrees of freedom.

```{r pval}
pval<-1-pchisq(3138.035, 1599)
pval
```
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(data=rw_data, aes(x=volatile.acidity))+
  geom_histogram(binwidth = 0.01, fill="lightblue", color="darkblue")+
  scale_x_continuous(breaks=seq(0.15,1,0.05), limits = c(0.125,1))+
  ggtitle("Volatile Acidity Counts")+
  xlab("Volatile Acidity (acetic acid) g/L")+
  ylab("Number of Samples")+
  theme(plot.title = element_text(hjust=0.5))+
  stat_density(geom="line",aes(y=0.01 * ..count..))
```

### Description One

The distribution of volatile acidity values is slightly bimodal.  The peaks are
around 0.40 g/L and 0.60 g/L.  There is a small third peak in the center around
0.50 g/L.  The mean is 0.53 g/L and the median is 0.52 g/L.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data=rw_data, aes(x=factor(quality), y=alcohol))+
  geom_boxplot()+
  ggtitle("Percieved Quality by Alcohol Content")+
  xlab("Quality (0-10)")+
  ylab("Percent Alcohol by Volume")+
  theme(plot.title = element_text(hjust=0.5))+
  scale_y_continuous(breaks = c(9,10,11,12,13,14,15))+
  geom_jitter(width = 0.35, alpha=1/5)
```

### Description Two

The samples with high alcohol content show an increase in perceived quality after
around 10% alcohol by volume.  Most values are found in the 5, 6, and 7 categories,
while comparatively fewer are found in the 3, 4, and 8 categories.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data=rw_data, aes(x=volatile.acidity/sulphates, y=alcohol))+
  ggtitle("Volatile Acidity/Sulphates (log_10) vs Alcohol")+
  xlab("Volatile Acidity/Sulphates (g/L)")+
  ylab("Percent Alcohol by Volume")+
  theme(plot.title = element_text(hjust=0.5))+
  geom_point(aes(colour=factor(quality)))+scale_color_brewer(palette='RdYlGn')+
  scale_x_log10(breaks=seq(0.15,4.00,0.6))+
  labs(color="Quality")
```

### Description Three

This plot uses a diverging color palette to show how low levels of volatile
acidity plus high levels of alcohol content and sulphates are associated with
better quality.  The presence of large number of mid-range quality values in the
bottom right region of the plot suggest that average quality can be achieved even
for samples with low alcohol content by maximizing the ratio of volatile acidity
to sulphates.

------

# Reflection

The red wine data set contains 1599 observations across 11 variables.  I made
histograms of each variable to understand their distributions, looked for
correlations between variables, and used box plots to check the effect of each
variable on quality.  Finally,  I used an Ordinal Logistic Regression with the
variables that most affected the wine quality as predictors.

The correlations pointed to volatile acidity, alcohol, sulphates, and citric acid
being the factors that most affected quality.  All the other factors had
correlations below 0.20.  Also, based on what I read in the information sheet
that came with the data, each of these factors have well understood mechanisms 
that affect the taste of red wine.  The box plots also supported the correlations
so I tried creating several models using combinations of these variables as
predictors.  Volatile acidity, alcohol, and sulphates together as predictors all
had p-values near zero, but the p-value of citric acid when it was added on was
around 0.20 and it didn't have any significant effect on the residual deviance.
Also, surprisingly, the correlation was negative in contradiction to the previously
found trend.  These issues led to citric acid being dropped from the model.

If this analysis were to be repeated, I would like there to be some improvements
made to the data.  The quality of each sample was said, in the information sheet,
to be based on the median evaluation of at least three experts.  I think it would
be better to increase the number of evaluations to each sample in order to minimize
bias. I would also like to see more data points in general than 1599.  For
example, the same study done using white wine had over 4,000 samples.  The model
in this analysis could be applied to this new data set to test its accuracy.

# Citation

[1] P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.